---
title: "MET500 - Data Import and Exploration"
author: "Charlie Whittaker"
date: "03/25/26"
output: 
  html_document:
    toc: true
    toc_depth: 4
  html_notebook: default
  pdf_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, cache=TRUE,
                      dev="png", message=FALSE, error=FALSE, warning=TRUE)
```

# Load Libraries

```{r}
library(openxlsx)
library(tidyverse)
library(ggrepel)
library(ggpubr)
library(XML)
library(reprex)
library(matrixStats)
library(readxl)

library(Seurat)
library(SeuratWrappers)
#library(Azimuth)
library(patchwork)
library(sctransform)
library(celldex)
library(SingleR)
library(rtracklayer)
library(apeglm)
library(ComplexHeatmap)
library(gprofiler2)
library(cluster)
library(fgsea)
library(SingleCellExperiment)
library(VAM)
library(UCell)
library(GSVA)
library(GSEABase)
library(ComplexHeatmap)
library(circlize)
library(grid)

#library(SoupX)
#library(DoubletFinder)
```

# Load Local Code

```{r}
source('Rcode/data_formatting_tools.R')
source('Rcode/analysis_tools.R')
source('Rcode/plots.R')
source('Rcode/common.R')
source('Rcode/ssGSEA.Library.R')
```

# Read in MET500 data

## Metadata

```{r}
meta <- read.table("../data/MET500_geneExpression_M.meta.plus.txt", header = TRUE, row.names = 1, sep = "\t")
```

## Gene expression data

Note, the l2 data provided seems to have been transformed without a +1 offset so this needs to be redone

```{r}
M.mx <- read.table("../data/MET500_geneExpression_M.mx.txt", header = TRUE, row.names = 1, sep = "\t")
# M.mx.l2 <- read.table("../data/MET500_geneExpression_M.mx.log2.txt", header = TRUE, row.names = 1, sep = "\t")
M.mx.l2.z <- read.table("../data/MET500_geneExpression_M.mx.log2.zscore.txt", header = TRUE, row.names = 1, sep = "\t")
```

```{r}
M.mx.l2 <- log2(M.mx+1)
```

## Read in annotation

```{r}
annot <- read.table("../data/probeMap_gencode.v23.annotation.gene.probemap", header = TRUE, row.names = 1, sep = "\t")
```

## Join gene from annot to gene expression

```{r}
# Step 1: Extract rownames as column
annot_df <- annot %>% rownames_to_column("gene_id") %>% dplyr::select(gene_id, gene)
expr_df <- as.data.frame(M.mx.l2) %>% rownames_to_column("gene_id")

# Step 2: Inner join (only rows in M.mx.l2)
M.mx.l2.annotated <- expr_df %>%
  inner_join(annot_df, by = "gene_id")

# Step 3: Move gene_id back to rownames
rownames(M.mx.l2.annotated) <- M.mx.l2.annotated$gene_id
M.mx.l2.annotated <- M.mx.l2.annotated %>% dplyr::select(-gene_id)

# Step 4: Report on join success
cat("=== JOIN REPORT ===\n\n")
cat("Rows in annot:              ", nrow(annot), "\n")
cat("Rows in M.mx.l2:            ", nrow(M.mx.l2), "\n")
cat("Rows after inner_join:      ", nrow(M.mx.l2.annotated), "\n\n")

# Check for unmatched
unmatched_annot <- anti_join(annot_df, expr_df, by = "gene_id")
unmatched_expr <- anti_join(expr_df, annot_df, by = "gene_id")

cat("Genes in annot but NOT in M.mx.l2:    ", nrow(unmatched_annot), "\n")
cat("Genes in M.mx.l2 but NOT in annot:    ", nrow(unmatched_expr), "\n\n")

# Success status
if (nrow(M.mx.l2.annotated) == nrow(M.mx.l2)) {
  cat("✓ SUCCESS: All rows from M.mx.l2 matched to annotation!\n")
} else {
  cat("⚠ WARNING: Some rows from M.mx.l2 did NOT match annotation\n")
  cat("  Unmatched genes:", paste(head(unmatched_expr$gene_id, 5), collapse=", "), "\n")
}

# View result
cat("\nFinal structure:\n")
print(head(M.mx.l2.annotated[, 1:8]))
```

## Repair data

After annotation, there are examples of multiple rows per gene, this needs to be repaired using the MPG approach

### Calculate means

```{r}
M.mx.l2.annotated <- mutate(M.mx.l2.annotated, Avg = rowMeans(as.matrix(dplyr::select(M.mx.l2.annotated, ends_with("XX")), na.rm = TRUE)))
```

### Adding MPG column

Often multiple geneids will map to the same gene symbol. In these cases, it is helpful to select a representative geneID for each group of gene symbols. The code block below iterates through each unique gene symbol, compares all of the geneids associated with the same gene symbol and identifies the geneid with the highest average expression. This gene is then set as the representative (max-per-gene) geneid for that symbol. Note: If multiple geneids have the same highest avg expression, the first gene that appears in the dataframe is taken

The code below performs the following steps:
  1) Create a vector called "mpg" of "No" values the same length as the number of rows in the dataframe assembleData
  2) Loop over the assembleDat object ... identifying index values of duplicate genes, updating mpg vector
  3) Merge mpg vector to assemblDat object with setting column name as "MPG"


```{r}
mpg <- rep('No', nrow(M.mx.l2.annotated))

for (i in unique(M.mx.l2.annotated[,"gene"])){
    if (is.na(i)){next}
    idxs <- which(M.mx.l2.annotated[,"gene"] == i)
    max_idx <- which(M.mx.l2.annotated$Avg[idxs] == max(M.mx.l2.annotated$Avg[idxs]))[1]
    
    mpg[idxs[max_idx]] <- 'Yes'
}

M.mx.l2.annotated$MPG <- mpg
```

### Select only MPG == Yes and remove Avg col

```{r}
M.mx.l2.annotated <- M.mx.l2.annotated %>% dplyr::filter(MPG == "Yes") %>% dplyr::select(-Avg,-MPG)
```

### Move gene to rowname and convert to matrix

```{r}
expr_matrix <- M.mx.l2.annotated %>%
  rownames_to_column("old_id") %>%
  dplyr::select(-old_id) %>%
  column_to_rownames("gene")

expr_matrix <- as.matrix(expr_matrix)
```

## Row-center and transpose the gene expression data for use later

```{r}
# Remove genes with zero variance BEFORE scaling
gene_sds <- apply(expr_matrix, 1, sd)
expr_matrix.filtered <- expr_matrix[gene_sds > 0, ]

cat("Genes removed (zero variance):", nrow(expr_matrix) - nrow(expr_matrix), "\n")
cat("Genes remaining:", nrow(expr_matrix), "\n")

# Now row-center
expr_matrix.rc <- calculate_zscore(expr_matrix.filtered)

# Verify - should have NO NaN
print(sum(is.nan(expr_matrix.rc)))
```

```{r}
expr_matrix.rc.transposed <- t(expr_matrix.rc)
```

# Gene Sets

```{r}
Ferroptosis <- c("CBS","PRNP","MAP1LC3B","SLC7A11","ACSL6","SAT2","TF","SLC11A2","CP","FTL","LPCAT3","ATG7","CTH",
                 "CYBB","AKR1C1","AKR1C2","DPP4","ACSL1","ACSL3","ACSL4","FDFT1","SLC39A14","ALOX15","FTH1","GCH1",
                 "NOX1","COQ2","GCLC","GCLM","GPX4","GSS","SLC40A1","HMGCR","HMOX1","HSPB1","IREB2","MAP1LC3C",
                 "NOX4","PCBP1","PCBP2","CHMP5","ACSL5","PHKG2","POR","STEAP3","CISD1","BACH1","SAT1","SLC39A8",
                 "SLC1A5","SLC3A2","TFRC","TP53","TXNRD1","VDAC2","VDAC3","CHMP6","NCOA4","SLC38A1","MAP1LC3A",
                 "AIFM2","AKR1C3","FTMT","ATG5")

Iron_intake <- c("SCARA5","LRP1","CD163","CYBRD1","FRRS1","SLC11A2","FLVCR2","SLC48A1","STEAP1","STEAP2","STEAP3",
                 "STEAP4","TF","TFR2","TFRC","SLC39A8","SLC39A14")

Iron_utilization <- c("FTH1","FTL","HMOX1","HIF1A","EPAS1","HIF3A","SLC25A37","SLC25A28","EGLN2","EGLN1","EGLN3","P4HTM")

Iron_efflux <- c("CP","SLC40A1","HAMP","FLVCR1","HEPH")

Fatty_acid <- c("ACAA1","ACAA2","ACACA","ACACB","ACAD10","ACAD11","ACAD8","ACAD9","ACADL","ACADM","ACADS","ACADSB","ACADVL","ACAT1",
        "ACAT2","ACBD3","ACBD4","ACBD5","ACBD6","ACBD7","ACOX1","ACOX2","ACOX3","ACOXL","ACSBG1","ACSBG2","ACSF2","ACSF3",
        "ACSL1","ACSL3","ACSL4","ACSL5","ACSL6","ACSM1","ACSM2A","ACSM2B","ACSM3","ACSM4","ACSM5","ACSS1","ACSS2","ACSS3",
        "CLYBL","CPT1A","CPT1B","CPT1C","CPT2","CYP4A11","CYP4A22","CYP4B1","CYP4F11","CYP4F12","CYP4F2","CYP4F22","CYP4F3",
        "CYP4F8","CYP4V2","CYP4X1","CYP4Z1","DCI","DECR1","DECR2","DGAT1","DGAT2","DGAT2L3","DGAT2L4","DGAT2L6","DGAT2L7",
        "ECH1","ECHDC1","ECHDC2","ECHDC3","ECHS1","EHHADH","ELOVL1","ELOVL2","ELOVL3","ELOVL4","ELOVL5","ELOVL6","ELOVL7",
        "ETFA","ETFB","ETFDH","FA2H","FAAH","FAAH2","FADS1","FADS2","FADS3","FAR1","FAR2","FASN","GPSN2","HACL1","HADH",
        "HADHA","HADHB","LIPC","LIPF","LIPG","LIPH","LIPI","LIPJ","LIPK","LIPM","LIPN","LPL","LPPR2","MCAT","MECR","MGLL",
        "MLYCD","MTTP","OLAH","OXSM","PECI","PECR","PHYH","PHYHD1","PNLIP","PNLIPRP1","PNLIPRP2","PNLIPRP3","PNPLA1",
        "PNPLA2","PNPLA3","PNPLA4","PNPLA5","PNPLA6","PNPLA7","PNPLA8","SCD","SCD5","SCP2","SDR16C6","SDR39U1","SDR42E1","SDR9C7")

Fatty_acid_uptake <- c("CD36","FABP4","SLC27A1","CAV1")

Unsaturated_FA_biosynthetic_process <- c("ABCD1","ABCD2","AKR1C3","ALOX12","ALOX15","ALOX15B","ANXA1","AVP","AVPR1A",
                                         "CBR1","CD74","DAGLB","DECR2","DEGS1","EDN1","EDN2","ELOVL1","ELOVL2","ELOVL3",
                                         "ELOVL4","ELOVL5","ELOVL6","ELOVL7","FABP5","FADS1","FADS2","FADS2B","FADS3",
                                         "HPGDS","IL1B","MIF","MIR132","MIR204","MIR766","PIBF1","PLA2G3","PLA2G4A","PLA2G4F",
                                         "PNPLA8","PRXL2B","PTGDS","PTGES","PTGES2","PTGES3","PTGIS","PTGS1","PTGS2","SCD",
                                         "SCD5","SIRT1","TBXAS1")

Membrane_Lipid <- c("AGPAT1","AGPAT2","AGPAT3","AGPAT4","AGPAT5","AGPAT6","AGPAT9","CDS1","CDS2","CEPT1","CHKA","CHKB","CHPT1",
                    "CLC","CRLS1","CROT","DAGLA","DAGLB","DGKA","DGKB","DGKD","DGKE","DGKG","DGKH","DGKI","DGKK","DGKQ","DGKZ",
                    "ETNK1","ETNK2","GDE1","GK","GK2","GK5","GNPAT","GPAM","GPD1","GPD1L","GPD2","LPCAT1","LPCAT2","LPCAT3",
                    "LPCAT4","LPGAT1","LPIN1","LPIN2","LPIN3","LYPLA1","LYPLA2","LYPLAL1","MBOAT1","MBOAT2","MBOAT4","MBOAT7",
                    "MOGAT1","MOGAT2","MOGAT3","NAPE-PLD","PAP2D","PCYT1A","PCYT1B","PCYT2","PEMT","PGS1","PHOSPHO1","PISD",
                    "PLA1A","PLA2G10","PLA2G12A","PLA2G12B","PLA2G15","PLA2G16","PLA2G1B","PLA2G2A","PLA2G2C","PLA2G2D","PLA2G2E",
                    "PLA2G2F","PLA2G3","PLA2G4A","PLA2G4B","PLA2G4C","PLA2G4D","PLA2G4E","PLA2G4F","PLA2G5","PLA2G6","PLA2G7",
                    "PLB1","PLCB1","PLCB2","PLCB3","PLCB4","PLCD1","PLCD3","PLCD4","PLCE1","PLCG1","PLCG2","PLCH1","PLCH2",
                    "PLCL1","PLCL2","PLCXD1","PLCXD2","PLCXD3","PLCZ1","PLD1","PLD2","PLD3","PLD4","PLD5","PLD6","PPAP2A",
                    "PPAP2B","PPAP2C","PPAPDC1A","PPAPDC1B","PPAPDC2","PPAPDC3","PRG2","PTDSS1","PTDSS2")

Redox <- c("CP","CRYZ","CRYZL1","CYB561","CYB5A","CYB5B","CYB5D1","CYB5D2","CYB5R1","CYB5R2","CYB5R3","CYB5R4","CYB5RL","CYBA",
           "CYBASC3","CYBB","CYBRD1","DHRS1","DHRS11","DHRS12","DHRS13","DHRS2","DHRS3","DHRS4","DHRS4L1","DHRS4L2","DHRS7",
           "DHRS7B","DHRS7C","DHRS9","DHRSX","DHTKD1","ENOX1","ENOX2","FDX1","FDX1L","FDXR","FOXRED1","FOXRED2","FRRS1","GLRX",
           "GLRX2","GLRX3","GLRX5","GRXCR1","GRXCR2","ISCA1","ISCA1L","ISCA2","LPO","MICAL1","MICAL2","MICAL3","NDOR1","NMRAL1",
           "OGFOD1","OGFOD2","OXA1L","PRDX1","PRDX2","PRDX3","PRDX4","PRDX5","PRDX6","PYROXD1")

# Ferroptosis_pathways2

Complex_I_IV <- c("MT-ND1","MT-ND2","MT-ND3","MT-ND4","MT-ND4L","MT-ND5","MT-ND6","NDUFA1","NDUFA10","NDUFA11","NDUFA12",
                  "NDUFA13","NDUFA2","NDUFA3","NDUFA4","NDUFA4L","NDUFA4L2","NDUFA5","NDUFA6","NDUFA7","NDUFA8","NDUFA9",
                  "NDUFAB1","NDUFAF1","NDUFAF2","NDUFB1","NDUFB10","NDUFB11","NDUFB2","NDUFB3","NDUFB4","NDUFB5","NDUFB6",
                  "NDUFB7","NDUFB8","NDUFB9","NDUFC1","NDUFC2","NDUFS1","NDUFS2","NDUFS3","NDUFS4","NDUFS5","NDUFS6","NDUFS7",
                  "NDUFS8","NDUFV1","NDUFV2","NDUFV3","SDHA","SDHB","SDHC","SDHD","CYC1","CYCS","MT-CYB","UCRC","UQCR",
                  "UQCRB","UQCRC1","UQCRC2","UQCRFS1","UQCRH","UQCRQ","COX1","COX10","COX11","COX15","COX16","COX17","COX18",
                  "COX19","COX2","COX3","COX4I1","COX4I2","COX5A","COX5B","COX6A1","COX6A2","COX6B1","COX6B2","COX6C","COX7A1",
                  "COX7A2","COX7A2L","COX7B","COX7B2","COX7C","COX8A","COX8C")

Reactive_oxygen_species_pathway <- c("ABCC1","ATOX1","CAT","CDKN2D","EGLN2","ERCC2","FES","FTL","G6PD","GCLC","GCLM","GLRX",
                                     "GLRX2","GPX3","GPX4","GSR","HHEX","HMOX2","IPCEF1","JUNB","LAMTOR5","LSP1","MBP","MGST1",
                                     "MPO","MSRA","NDUFA6","NDUFB4","NDUFS2","NQO1","OXSR1","PDLIM1","PFKP","PRDX1","PRDX2",
                                     "PRDX4","PRDX6","PRNP","PTPA","SBNO2","SCAF4","SELENOS","SOD1","SOD2","SRXN1","STK25",
                                     "TXN","TXNRD1","TXNRD2")

LOXs <- c("ALOX12","ALOX12B","ALOX15","ALOX15B","ALOX5","ALOX5AP","ALOX8","ALOXE3")

BH4 <- c("GCH1","PTS","QDPR","SPR")

Folate <- c("DHFR","DHFRL1","FPGS","MTHFD1","MTHFD1L","MTHFD2","MTHFD2L","MTHFR","MTHFS","MTHFSD","SHMT1","SHMT2")

Glutathione <- c("GCLC","GCLM","GPX1","GPX2","GPX3","GPX4","GPX5","GPX6","GPX7","GPX8","GSR","GSS","GSTA1","GSTA2","GSTA3",
                 "GSTA4","GSTA5","GSTCD","GSTK1","GSTM1","GSTM2","GSTM3","GSTM4","GSTM5","GSTO1","GSTO2","GSTP1","GSTT1",
                 "GSTT2","HAGH","HAGHL","MGST1","MGST2","MGST3")

CoQ10 <- c("PDSS1","PDSS2","COQ2","COQ3","COQ6","COQ5","COQ7","COQ8A","COQ9","COQ8B")

Thioredoxin <- c("TXNRD1","TXNRD2","TXNRD3","NXN","TXN","SELENOT","TXNDC2","PRDX1","PRDX2","PRDX3","PRDX4","PRDX5")

Glutamate_transmembrane <- c("ARL6IP1","ARL6IP5","ATP1A2","CLN8","GRM1","ITGB1","KCNJ10","NTSR1","PER2","PRAF2","PSEN1",
                             "SLC17A6","SLC17A7","SLC17A8","SLC1A1","SLC1A2","SLC1A3","SLC1A6","SLC1A7","SLC25A12","SLC25A13",
                             "SLC25A18","SLC25A22","SLC38A7","SLC7A11")

homocysteine_metabolic_process <- c("BHMT","BLMH","CBS","CBSL","CPS1","CTH","DPEP1","MMUT","MPST","MTHFR","MTRR","NOX4","TST")

Pentose_Phosphate <- c("DERA","G6PD","PGD","PGLS","RPE","RPIA","TALDO1","TKT","TKTL1","TKTL2")

Glycolysis <- c("ADPGK","ALDOA","ALDOB","ALDOC","BPGM","DLAT","ENO1","ENO2","ENO3","GAPDH","GAPDHS","GCK","GPI","HK1","HK2",
                "HK3","HKDC1","PDHA1","PDHA2","PDHB","PDHX","PFKFB1","PFKFB2","PFKFB3","PFKFB4","PFKL","PFKM","PFKP","PGAM1",
                "PGAM2","PGAM4","PGAM5","PGK1","PGK2","PGM1","PGM2","PGM2L1","PGM5","PKLR","PKM2","TPI1")

Krebs <- c("ACLY","ACO1","ACO2","CS","D2HGDH","DLST","FH","HCCS","IDH1","IDH2","IDH3A","IDH3B","IDH3G","L2HGDH","MDH1","MDH1B",
           "MDH2","ME1","ME2","ME3","OGDH","OGDHL","PC","SUCLA2","SUCLG1","SUCLG2")

# EMT and Others

EMT.Klughammer <- c("IGFBP7","SPARC","COL1A2","COL4A1","COL3A1","BGN","ACTA2","FN1","COL4A2","TAGLN",
                    "DCN","COL1A1","LUM","COL6A3","POSTN","AEBP1","COL6A2","VIM","TIMP1","TPM2","COL5A1",
                    "CALD1","COL6A1","A2M","SPARCL1","THY1","VCAN","CCN2","GNG11","PDGFRB","RGS5",
                    "ITGA1","MYL9","COL5A2","COL18A1","THBS2","IGHA1","CAVIN1","ELN","NID1","LHFPL6",
                    "APOE","IGLC3","HSPG2","CAV1","TCF4","NNMT","ASPN","FSTL1","MGP")

EMT.Hallmark <- c("ABI3BP", "ACTA2","ADAM12","ANPEP","APLP1","AREG","BASP1","BDNF","BGN","BMP1",
                  "CADM1","CALD1","CALU","CAP2","CAPG","CCN1","CCN2","CD44","CD59","CDH11","CDH2",
                  "CDH6","COL11A1","COL12A1","COL16A1","COL1A1","COL1A2","COL3A1","COL4A1","COL4A2",
                  "COL5A1","COL5A2","COL5A3","COL6A2","COL6A3","COL7A1","COL8A2","COLGALT1","COMP",
                  "COPA","CRLF1","CTHRC1","CXCL1","CXCL12","CXCL6","CXCL8","DAB2","DCN","DKK1","DPYSL3",
                  "DST","ECM1","ECM2","EDIL3","EFEMP2","ELN","EMP3","ENO2","FAP","FAS","FBLN1","FBLN2",
                  "FBLN5","FBN1","FBN2","FERMT2","FGF2","FLNA","FMOD","FN1","FOXC2","FSTL1","FSTL3",
                  "FUCA1","FZD8","GADD45A","GADD45B","GAS1","GEM","GJA1","GLIPR1","GPC1","GPX7",
                  "GREM1","HTRA1","ID2","IGFBP2","IGFBP3","IGFBP4","IL15","IL32","IL6","INHBA","ITGA2",
                  "ITGA5","ITGAV","ITGB1","ITGB3","ITGB5","JUN","LAMA1","LAMA2","LAMA3","LAMC1","LAMC2",
                  "LGALS1","LOX","LOXL1","LOXL2","LRP1","LRRC15","LUM","MAGEE1","MATN2","MATN3","MCM7",
                  "MEST","MFAP5","MGP","MMP1","MMP14","MMP2","MMP3","MSX1","MXRA5","MYL9","MYLK","NID2",
                  "NNMT","NOTCH2","NT5E","NTM","OXTR","P3H1","PCOLCE","PCOLCE2","PDGFRB","PDLIM4","PFN2",
                  "PLAUR","PLOD1","PLOD2","PLOD3","PMEPA1","PMP22","POSTN","PPIB","PRRX1","PRSS2","PTHLH",
                  "PTX3","PVR","QSOX1","RGS4","RHOB","SAT1","SCG2","SDC1","SDC4","SERPINE1","SERPINE2",
                  "SERPINH1","SFRP1","SFRP4","SGCB","SGCD","SGCG","SLC6A8","SLIT2","SLIT3","SNAI2",
                  "SNTB1","SPARC","SPOCK1","SPP1","TAGLN","TFPI2","TGFB1","TGFBI","TGFBR3","TGM2",
                  "THBS1","THBS2","THY1","TIMP1","TIMP3","TNC","TNFAIP3","TNFRSF11B","TNFRSF12A",
                  "TPM1","TPM2","TPM4","VCAM1","VCAN","VEGFA","VEGFC","VIM","WIPF1","WNT5A")

EMT.prelim <- c("VIM","ZEB1","ZEB2","ITGA2")

EMT.76gs <- c("CHEK2","FZR1","CHAF1B","CDS1","AP1M2","FN1","EPPK1","BSPRY","HNMT","PRR5","ST14","EPN3","ZEB1",
              "CLDN7","SH3YL1","PRSS22","TGFBI","CRB3","MUC1","DSP","RBPMS","ELMO3","CLDN4","ENPP5","EPB41L5",
              "MPZL2","EPHA1","PPARG","RAB25","CARD6","SCNN1A","ERBB3","SPINT2","EVPL","TACSTD2","TMC4","CDH1",
              "TNFRSF21","FXYD3","MPP7","GALNT3","GALNT5","NRP1","AXL","PRSS8","GRHL1","GRHL2","S100A14","CDH3",
              "SERINC2","SHROOM3","ITGB6","SSH3","KLC3","STAP2","ANTXR2","ANKRD22","TJP3","LIX1L","TMEM125",
              "TMEM45B","MAL2","VIM","MAPK13","MMP2","KRT19","KRTCAP3","ESRP1","XXYLT1","TC2N","ADGRF1","ADGRG1",
              "S100A11P1","CDH15","PATJ","TSKU")

EMT.MLR_method <- c("FZR1","PAPPA-AS1","CCM2","CDKN2AIP","CLDN7","VIM","CDH1","GRHL2","OVOL1","SLC25A42","SNX13",
                   "TAF4B","CRIP2","CDK2","MBNL1","NEURL1B","ANG","RNASE4","PPFIBP1","PACSIN1","LRRTM1","TMEM182",
                   "CSMD1","ZNF503-AS2","CCNF","DIRC1","MBTPS2","RNF150","RC3H2","UBE3C","HUNK","CEP135","SAFB2",
                   "OSM","MTAP","PDS5B","ACTN4","AFF3","THAP4","ZNF490","CLU","GGCX","CARF","REXO2","LIG3",
                   "SRD5A3-AS1","FZD10-AS1","ADAMTS7P3","ADAMTS7","MIAT","COG1","CD47","PLEKHB2","BAIAP2-DT")

EMT.cell_line.epi <- c("CDH1","AGR2","EPCAM","KRT19","RAB25","TACSTD2","S100P","CEACAM6","GALNT3","FXYD3","SPINT2",
                       "TMEM30B","SCNN1A","ST14","ESRP1","S100A14","CLDN7","ERBB3","RBM47","SPINT1","ELF3","CLDN4",
                       "PRSS8","SH3YL1","EHF","LCN2","JUP","VAMP8","KRT8","C1orf106","KRT7","DSP","CDS1","ITGB4",
                       "TMPRSS4","LSR","SORL1","GRHL2","PPL","C1orf116","TSPAN1","MAP7","SLPI","TOX3","ARHGAP8","F11R",
                       "LAD1","GPX2","CTSH","GPR56","FA2H","KLF5","AREG","KRT18","SCEL","CDH3","UGT1A1","MPZL2","AIM1",
                       "OVOL2","LLGL2","ESRP2","MYO5C","DDR1","VGLL1","IRF6","SFN","TSPAN13","KCNK1","MYO1D","PKP3",
                       "ITGB6","LY75","MAPK13","TTC39A","ELMO3","CEACAM1","DTX4","ERBB2","RAB11FIP1","ATP2C2","MST1R",
                       "AP1M2","TGFA","MYO6","PTK6","OAS1","FBP1","AQP3","CBLC","EPHA1","BSPRY","SH2D3A","EPS8L1","GRB7",
                       "C4orf19","KLK6","TJP2","PLS1","DENND2D","EPS8L2","IL20RA","HES1","IL1RN","EXPH5","ARHGDIB",
                       "C19orf21","CAMK2N1","HPGD","SYNGR2","C10orf116","PERP","MANSC1","DSC2","POF1B","SERINC5","BIK",
                       "ANXA9","MALL","EPN3","STAP2","FOXA1","PYCARD","ZNF165","SLC37A1","ANK3","TSPAN15","HNMT","ABCC3",
                       "SDC1","CKMT1A","TOB1","B3GNT3","TMC6","CD9","ADAP1","ATP1B1","SHANK2","CYB561","ERMP1","RAB20",
                       "MYH14","CAPN1","ALDH3B2","TRIM31","ARAP2","SSH3","ICA1","ARHGEF5","ALOX5","RHOD","TMPRSS2","MTUS1",
                       "CYP4F3","PPFIBP2","RABGAP1L","PLXNB2","MGST2","OR7E14P","EVPL","CD46","KRT15","CNKSR1","BLNK",
                       "COMT","ANXA4","TNFSF13","OCLN","SLC9A3R1","XBP1")

EMT.cell_line.mes <- c("VIM","ZEB1","EMP3","SACS","AXL","LOXL2","SPARC","FHL1","FERMT2","TUBA1A","TMEM158","CALD1",
                       "LGALS1","PMP22","MSN","GLYR1","MAP1B","AP1S2","GJA1","DENND5A","C12orf24","TPM2","TUBB6",
                       "SRPX","ANK2","CHN1","SH2B3","LEPRE1","ETV1","SOBP","AKAP12","TGFB1I1","SERPINE1","SOAT1",
                       "COL5A2","LHFP","CEP170","POPDC3","TRPC1","KDELC1","MYL9","BAG2","FSTL1","MXRA7","GFPT2",
                       "RECK","TMEFF1","PTRF")

EMT.tumor.epi <- c("KRT19","AGR2","RAB25","CDH1","ERBB3","FXYD3","SLC44A4","S100P","SCNN1A","GALNT3","PRSS8","ELF3",
                   "CEACAM6","TMPRSS4","CLDN7","TACSTD2","CLDN3","EPCAM","SPINT1","TSPAN1","PLS1","TMEM30B","PRR15L",
                   "KRT8","ST14","RBM47","S100A14","C1orf106","NQO1","TOX3","PTK6","TFF1","CLDN4","GPRC5A","TJP3",
                   "KRT18","MAP7","CKMT1A","ESRP1","MUC1","SPINT2","ESRP2","CDS1","PPAP2C","CEACAM7","TTC39A","OVOL2",
                   "EHF","AP1M2","CEACAM5","LAD1","ARHGAP8","TFF3","JUP","CD24","TMC5","MLPH","ELMO3","ERBB2","LLGL2",
                   "DDR1","FA2H","CBLC","TMPRSS2","LSR","PERP","POF1B","MYO5C","RAB11FIP1","MAPK13","KRT7","CEACAM1",
                   "CXADR","ATP2C2","RNF128","MPZL2","EPS8L1","GALNT7","CORO2A","BCAS1","TPD52","ARHGAP32","FUT2",
                   "OR7E14P","GALE","GRHL2","BIK","RAPGEFL1","STYK1","F11R","PKP3","CYB561","SH3YL1","GDF15","PSCA",
                   "EZR","TJP2","FGFR3","FUT3","BSPRY","TOM1L1  ","IRF6","EPB41L4B","OCLN","LRRC1","C19orf21",
                   "ABHD11","EPS8L2","MYO6","TSPAN8","MST1R","SLC16A5","GPR56","AZGP1","TOB1","SLC35A3","TRPM4",
                   "PHLDA2","VAMP8","SLC22A18","AKR1B10","VAV3","SPAG1","ABCC3","SYNGR2","STAP2","C4orf19","PPL",
                   "PLLP","DSG2","HDHD3","CD2AP","MANSC1","DHCR24","EPN3","TUFT1","GMDS","EXPH5","DSP","SDC4","IL20RA",
                   "FAM174B","PTPRF","SORD")

EMT.tumor.mes <- c("GAS1","CXCL12","ZEB1","GLYR1","FHL1","FERMT2","C1S","FYN","WIPF1","CYP1B1","SERPING1","SERPINF1",
                   "VCAM1","MAP1B","TCF4","SRPX","EMP3","DPT","CALD1","PTGIS","VIM","CD163","C1R","FBN1","FN1","FXYD6",
                   "IGF1","NAP1L3","MRC1","QKI","MS4A4A","DCN","LOX","RECK","ANK2","LY96","ZFPM2","CSRP2","EFEMP1",
                   "RARRES2","PTPRC","PLEKHO1","RGS2","F13A1","JAM2","CHRDL1","TUBA1A","AP1S2","MYLK","DDR2","DSE",
                   "SACS","GLIPR1","CXCL13","FLRT2","PTX3","AKT3","COL6A2","DPYSL3","CDH11","PDZRN3","ZEB2","CCL2",
                   "MAFB","SFRP1","C14orf139","MFAP4","MAF","UCHL1","TUBB6","SRGN","HEG1","KCNJ8","AKAP12","EVI2A",
                   "COL14A1","AXL","ECM2","FSTL1","PLN","MYL9","OLFML3","STON1","SLIT2","BICC1","SOBP","CLIC4","ENPP2",
                   "SAMSN1","TPM2","ASPN","COL6A1","IGFBP5","MOXD1","AKAP2","SLC2A3","OLFML2B","ANGPTL2","PCOLCE",
                   "COLEC12","CTSK","TAGLN","CDH2","IL10RA","C1orf54","CEP170","TNS1","CLEC2B","JAM3","SEPT6","GREM1",
                   "VCAN","ZCCHC24","CRYAB","SFRP4","RUNX1T1","FGL2","MS4A6A","PTRF","GIMAP4","TWIST1","GFPT2","LHFP",
                   "CXCR4","SPOCK1","SPARC","VSIG4","GPM6B","TRPC1","SNAI2","GUCY1B3","PLXNC1","SYT11","FLI1","MYH10",
                   "CSF2RB","TNC","PMP22","COL5A2","MMP2","GNG11","CAV1","CDK14","SDC2","PTGDS","NR3C1","SYNM","FAP",
                   "NUAK1","WWTR1","FBLN1","MPDZ","SYNE1","EFEMP2","GIMAP6","KIAA1462","CCL8","COL15A1","CHN1","CRISPLD2",
                   "PDGFC","GEM","ISLR","GZMK","SPARCL1","BNC2","BGN","MEOX2","ITM2A","IFFO1")

LAM <- c("TREM2","APOE","C1Q2A","C1QB","C1QC","CD9","CD163")

exhausted.CD8 <- c("PDCD1","CTLA4","LAG3","HAVCR2","TIGIT","TOX","EOMES","TBX21","CXCL13","CD244")
```

## Score gene sets for heatmap using gsva

the heatmap contains the following gene sets:

Gene list       pathway type
Fatty acid uptake       Ferroptosis promotion
LOXs    Ferroptosis promotion
Iron intake WHv1        Ferroptosis promotion
Iron utilization        Ferroptosis promotion
Glutathione     Ferroptosis suppression
BH4     Ferroptosis suppression
CoQ10   Ferroptosis suppression
ACSL4   Gene
AGPS    Gene
FAR1    Gene

```{r}
gsva_params <- gsvaParam(exprData = expr_matrix, 
                         geneSets = list(
                           Fatty_acid_uptake.gsva = Fatty_acid_uptake,
                           LOXs.gsva = LOXs,
                           Thioredoxin.gsva = Thioredoxin,
                           Glutamate_transmembrane.gsva = Glutamate_transmembrane,
                           Membrane_Lipid.gsva = Membrane_Lipid,
                           Unsat_FA_biosynth_process.gsva = Unsaturated_FA_biosynthetic_process,
                           Fatty_acid.gsva = Fatty_acid,
                           Iron_intake.gsva = Iron_intake,
                           Iron_utilization.gsva = Iron_utilization,
                           Glutathione.gsva = Glutathione,
                           BH4.gsva = BH4,
                           CoQ10.gsva = CoQ10
                         ))

Heatmap.GSVA.scores <- gsva(gsva_params)
```

### Transpose gsva scores

```{r}
Heatmap.GSVA.scores.transposed <- t(Heatmap.GSVA.scores)
```

# EMT scoring with Chakraborty KS methd

## Import KS code

```{r}
source("./emt_code/Chakraborty_et_al_KS_code_vCAW.R")
```

## Point to signature file

```{r}
KSScore.EMTSignature.file = "./emt_code/EM_gene_signature_tumor_KS.xlsx"
```

## Calculate KS

```{r}
met500.ks.emt <- KSScore(expr_matrix,"met500",".")
met500.ks.emt <- as.data.frame(met500.ks.emt)
```

# Combine data for plotting

```{r}
# Convert all to dataframes with rownames as column
gsva_df <- as.data.frame(Heatmap.GSVA.scores.transposed) %>% 
  rownames_to_column("Sample")
expr_df <- as.data.frame(expr_matrix.rc.transposed) %>% 
  rownames_to_column("Sample")
met_df <- as.data.frame(met500.ks.emt) %>% 
  rownames_to_column("Sample")

# Join all three by Sample
merged_all <- gsva_df %>%
  inner_join(expr_df, by = "Sample") %>%
  inner_join(met_df, by = "Sample")

# Convert Sample back to rownames
rownames(merged_all) <- merged_all$Sample
merged_all <- merged_all %>% dplyr::select(-Sample)

# Report on join success
cat("=== THREE-WAY JOIN REPORT ===\n\n")
cat("Rows in GSVA scores:                  ", nrow(gsva_df), "\n")
cat("Rows in expression matrix:            ", nrow(expr_df), "\n")
cat("Rows in met500.ks.emt:                ", nrow(met_df), "\n")
cat("Rows after inner_join:                ", nrow(merged_all), "\n\n")

# Check for unmatched
unmatched_gsva <- anti_join(gsva_df, expr_df, by = "Sample") %>% 
  anti_join(met_df, by = "Sample")
unmatched_expr <- anti_join(expr_df, gsva_df, by = "Sample") %>% 
  anti_join(met_df, by = "Sample")
unmatched_met <- anti_join(met_df, gsva_df, by = "Sample") %>% 
  anti_join(expr_df, by = "Sample")

cat("Samples in GSVA but NOT in others:    ", nrow(unmatched_gsva), "\n")
cat("Samples in expr but NOT in others:    ", nrow(unmatched_expr), "\n")
cat("Samples in met but NOT in others:     ", nrow(unmatched_met), "\n\n")

# Success status
if (nrow(merged_all) > 0) {
  cat("✓ SUCCESS: Join completed!\n")
  cat("  Final dimensions: ", nrow(merged_all), " samples × ", ncol(merged_all), " variables\n\n")
} else {
  cat("✗ ERROR: No matching samples found!\n")
}

# View result
cat("Final structure:\n")
print(head(merged_all[, 1:8]))
```

## Gene set scatter plots

```{r,fig.height=6,fig.width=7}
# Helper function to get correlation statistics
get_cor_stats <- function(x, y) {
  # Remove NA values
  idx <- !is.na(x) & !is.na(y)
  x <- x[idx]
  y <- y[idx]
  
  # Calculate correlation
  cor_test <- cor.test(x, y, method = "pearson")
  r_squared <- cor_test$estimate^2
  p_value <- cor_test$p.value
  
  return(list(r_squared = r_squared, p_value = p_value))
}

# Plot 1: BH4.gsva
stats1 <- get_cor_stats(merged_all$met500.ks.emt, merged_all$BH4.gsva)
title1 <- sprintf("Scatterplot of met500.ks.emt vs BH4.gsva - (R² = %.2f, p = %.2e)", 
                  stats1$r_squared, stats1$p_value)

ggplot(merged_all, aes(x = met500.ks.emt, y = BH4.gsva)) +
  geom_point(color = "steelblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "red", alpha = 0.2) +
  labs(
    title = title1,
    x = "met500.ks.emt",
    y = "BH4 gsva"
  ) +
  theme_minimal()

# Plot 2: ACSL4
stats2 <- get_cor_stats(merged_all$met500.ks.emt, merged_all$ACSL4)
title2 <- sprintf("Scatterplot of met500.ks.emt vs ACSL4 - (R² = %.2f, p = %.2e)", 
                  stats2$r_squared, stats2$p_value)

ggplot(merged_all, aes(x = met500.ks.emt, y = ACSL4)) +
  geom_point(color = "steelblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "red", alpha = 0.2) +
  labs(
    title = title2,
    x = "met500.ks.emt",
    y = "ACSL4 (Z-Score)"
  ) +
  theme_minimal()

# Plot 3: GCH1
stats3 <- get_cor_stats(merged_all$met500.ks.emt, merged_all$GCH1)
title3 <- sprintf("Scatterplot of met500.ks.emt vs GCH1 - (R² = %.2f, p = %.2e)", 
                  stats3$r_squared, stats3$p_value)

ggplot(merged_all, aes(x = met500.ks.emt, y = GCH1)) +
  geom_point(color = "steelblue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "red", fill = "red", alpha = 0.2) +
  labs(
    title = title3,
    x = "met500.ks.emt",
    y = "GCH1 (Z-Score)"
  ) +
  theme_minimal()
```

# Prepare Heatmap version 1

## Select data to plot

```{r}
merged_all.df <- merged_all %>% rownames_to_column("Sample")

heat.data <- merged_all.df %>% 
  dplyr::select(all_of(c("Sample",
                         "met500.ks.emt",
                         "Fatty_acid_uptake.gsva","LOXs.gsva", "Iron_intake.gsva", "Membrane_Lipid.gsva", "Unsat_FA_biosynth_process.gsva", "Fatty_acid.gsva",
                         "Iron_utilization.gsva", "Glutathione.gsva", "BH4.gsva", "Thioredoxin.gsva","Glutamate_transmembrane.gsva","CoQ10.gsva",
                         "ACSL4","AGPS","FAR1","GCH1")))
```

```{r,fig.width=22,fig.height=7}
# Sort by EMT_KS_Score
heat.data.sorted <- heat.data[order(heat.data$met500.ks.emt), ]

# Define genes in order
genes_to_plot <- c("met500.ks.emt",
                   "Fatty_acid_uptake.gsva", "LOXs.gsva", "Iron_intake.gsva", "Membrane_Lipid.gsva", "Unsat_FA_biosynth_process.gsva", "Fatty_acid.gsva", 
                   "Iron_utilization.gsva", "Glutathione.gsva", "BH4.gsva","Thioredoxin.gsva","Glutamate_transmembrane.gsva","CoQ10.gsva", 
                   "ACSL4", "AGPS", "FAR1", "GCH1")

# Create matrix
heatmap_data <- as.matrix(heat.data.sorted[, genes_to_plot])
heatmap_data <- t(heatmap_data)

# Define groups
group_vec <- c("SortOrder",
               "Ferroptosis promotion", "Ferroptosis promotion", "Ferroptosis promotion","Ferroptosis promotion","Ferroptosis promotion","Ferroptosis promotion","Ferroptosis promotion",
               "Ferroptosis suppression", "Ferroptosis suppression","Ferroptosis suppression", "Ferroptosis suppression","Ferroptosis suppression",
               "Gene", "Gene", "Gene", "Gene"
)

# Create row split factor for spacing
row_split <- factor(group_vec, 
                    levels = c("SortOrder", "Ferroptosis promotion", 
                              "Ferroptosis suppression", "Gene"))

# Create color functions
# Row 1: SortOrder
# Rows 2-13: Ferroptosis (12 rows)
# Rows 14-17: Genes (4 rows)
ferroptosis_data <- heatmap_data[2:13, ]
gene_data <- heatmap_data[14:17, ]
sortorder_data <- heatmap_data[1, ]

# Ferroptosis color scale
ferroptosis_cols <- colorRamp2(c(min(ferroptosis_data), 0, max(ferroptosis_data)), 
                                c("blue", "white", "red"))

# Gene color scale: white to dark purple, with 75th percentile as cutoff
# gene_75_percentile <- quantile(gene_data, 0.75)
gene_cols <- colorRamp2(c(min(gene_data), 0, max(gene_data)), c("blue", "white", "red"))

# SortOrder color scale: orange to green
sortorder_cols <- colorRamp2(c(min(sortorder_data), max(sortorder_data)), 
                              c("darkorange", "lightgreen"))

# Step 6: Create heatmap WITH left annotation for group bars
ht <- Heatmap(heatmap_data,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_split = row_split,
        row_gap = unit(3, "mm"),
        row_names_side = "right",
        show_column_names = FALSE,
        show_heatmap_legend = FALSE,
        left_annotation = rowAnnotation(
          Group = group_vec,
          col = list(Group = c("SortOrder" = "darkgreen",
                               "Ferroptosis promotion" = "steelblue",
                               "Ferroptosis suppression" = "coral",
                               "Gene" = "brown")),
          show_legend = FALSE,
          simple_anno_size = unit(0.8, "cm"),
          gp = gpar(fontsize = 9)
        ),
        cell_fun = function(j, i, x, y, w, h, fill) {
          if(i == 1) {
            # SortOrder row
            fill_color <- sortorder_cols(heatmap_data[i, j])
          } else if(i <= 13) {
            # Ferroptosis rows
            fill_color <- ferroptosis_cols(heatmap_data[i, j])
          } else {
            # Gene rows
            fill_color <- gene_cols(heatmap_data[i, j])
          }
          grid.rect(x, y, w, h, gp = gpar(fill = fill_color, col = NA))
        },
        rect_gp = gpar(type = "none"),
        name = "Value",
        width = ncol(heatmap_data)*unit(0.3, "mm"),
        height = nrow(heatmap_data)*unit(9, "mm"))

# Create separate legends
group_legend <- Legend(at = c("SortOrder", "Ferroptosis promotion", 
                              "Ferroptosis suppression", "Gene"),
                       labels = c("SortOrder", "Ferroptosis promotion", 
                                  "Ferroptosis suppression", "Gene"),
                       legend_gp = gpar(fill = c("darkgreen", "steelblue", "coral", "brown")),
                       title = "Group Category")

gene_legend <- Legend(col_fun = gene_cols,
                      title = "Gene Expression\nZ-Score",
                      direction = "vertical")

ferroptosis_legend <- Legend(col_fun = ferroptosis_cols,
                             title = "GSVA Scores",
                             direction = "vertical")

sortorder_legend <- Legend(col_fun = sortorder_cols,
                           title = "Sort Order\n(EMT_KS_Score)",
                           direction = "vertical")

# Step 8: Draw with legends
pdf("heatmap.pdf", width = 22, height = 7)
draw(ht, 
     annotation_legend_list = list(group_legend, sortorder_legend, ferroptosis_legend, gene_legend),
     annotation_legend_side = "right",
     padding = unit(c(1, 1, 1, 1), "cm"))
dev.off()

# Or view in RStudio
draw(ht, 
     annotation_legend_list = list(group_legend, sortorder_legend, ferroptosis_legend, gene_legend),
     annotation_legend_side = "right",
     padding = unit(c(1, 1, 1, 1), "cm"))
```

# Prepare Heatmap version 2

This version removes the genes from the plot. removes LOX and CoQ10, and reorders the gene sets

```{r}
merged_all.df <- merged_all %>% rownames_to_column("Sample")

heat.data <- merged_all.df %>% 
  dplyr::select(all_of(c("Sample",
                         "met500.ks.emt",
                         "Fatty_acid_uptake.gsva","Iron_intake.gsva", "Membrane_Lipid.gsva", "Unsat_FA_biosynth_process.gsva", "Fatty_acid.gsva",
                         "Iron_utilization.gsva", "Glutathione.gsva", 
                         "BH4.gsva", "Thioredoxin.gsva","Glutamate_transmembrane.gsva")))
```

```{r,fig.width=22,fig.height=7}
# Sort by EMT_KS_Score
heat.data.sorted <- heat.data[order(heat.data$met500.ks.emt), ]

# Define genes in order
genes_to_plot <- c("met500.ks.emt",
                   "Fatty_acid_uptake.gsva", "Unsat_FA_biosynth_process.gsva", "Fatty_acid.gsva", "Membrane_Lipid.gsva", "Iron_intake.gsva", "Iron_utilization.gsva",
                   "BH4.gsva","Thioredoxin.gsva","Glutathione.gsva","Glutamate_transmembrane.gsva")

# Create matrix
heatmap_data <- as.matrix(heat.data.sorted[, genes_to_plot])
heatmap_data <- t(heatmap_data)

# Define groups
group_vec <- c("SortOrder",
               "Ferroptosis promotion", "Ferroptosis promotion", "Ferroptosis promotion","Ferroptosis promotion","Ferroptosis promotion","Ferroptosis promotion",
               "Ferroptosis suppression", "Ferroptosis suppression","Ferroptosis suppression", "Ferroptosis suppression")

# Create row split factor for spacing
row_split <- factor(group_vec, 
                    levels = c("SortOrder", 
                               "Ferroptosis promotion",
                               "Ferroptosis suppression"))

# Create color functions
# Row 1: SortOrder
# Rows 2-11: Ferroptosis (12 rows)
ferroptosis_data <- heatmap_data[2:11, ]
sortorder_data <- heatmap_data[1, ]

# Ferroptosis color scale
ferroptosis_cols <- colorRamp2(c(min(ferroptosis_data), 0, max(ferroptosis_data)), 
                                c("blue", "white", "red"))

# SortOrder color scale: orange to green
sortorder_cols <- colorRamp2(c(min(sortorder_data), max(sortorder_data)), 
                              c("darkorange", "lightgreen"))

# Create heatmap WITH left annotation for group bars
ht <- Heatmap(heatmap_data,
        cluster_rows = FALSE,
        cluster_columns = FALSE,
        row_split = row_split,
        row_gap = unit(3, "mm"),
        row_names_side = "right",
        show_column_names = FALSE,
        show_heatmap_legend = FALSE,
        left_annotation = rowAnnotation(
          Group = group_vec,
          col = list(Group = c("SortOrder" = "darkgreen",
                               "Ferroptosis promotion" = "steelblue",
                               "Ferroptosis suppression" = "coral")),
          show_legend = FALSE,
          simple_anno_size = unit(0.8, "cm"),
          gp = gpar(fontsize = 9)
        ),
        cell_fun = function(j, i, x, y, w, h, fill) {
          if(i == 1) {
            # SortOrder row
            fill_color <- sortorder_cols(heatmap_data[i, j])
          } else if(i <= 11) {
            # Ferroptosis rows
            fill_color <- ferroptosis_cols(heatmap_data[i, j])
          }
          grid.rect(x, y, w, h, gp = gpar(fill = fill_color, col = NA))
        },
        rect_gp = gpar(type = "none"),
        name = "Value",
        width = ncol(heatmap_data)*unit(0.3, "mm"),
        height = nrow(heatmap_data)*unit(9, "mm"))

# Create separate legends
group_legend <- Legend(at = c("SortOrder", "Ferroptosis promotion", 
                              "Ferroptosis suppression"),
                       labels = c("SortOrder", "Ferroptosis promotion", 
                                  "Ferroptosis suppression"),
                       legend_gp = gpar(fill = c("darkgreen", "steelblue", "coral")),
                       title = "Group Category")

ferroptosis_legend <- Legend(col_fun = ferroptosis_cols,
                             title = "GSVA Scores",
                             direction = "vertical")

sortorder_legend <- Legend(col_fun = sortorder_cols,
                           title = "Sort Order\n(EMT_KS_Score)",
                           direction = "vertical")

# Step 8: Draw with legends
pdf("heatmap_v2.pdf", width = 22, height = 7)
draw(ht, 
     annotation_legend_list = list(group_legend, sortorder_legend, ferroptosis_legend),
     annotation_legend_side = "right",
     padding = unit(c(1, 1, 1, 1), "cm"))
dev.off()

# Or view in RStudio
draw(ht, 
     annotation_legend_list = list(group_legend, sortorder_legend, ferroptosis_legend),
     annotation_legend_side = "right",
     padding = unit(c(1, 1, 1, 1), "cm"))
```

# write session info

Capturing information about the R session helps document your work

```{r}
sessionInfo()
writeLines(capture.output(sessionInfo()), "01_MET500_dataImport_sessionInfo.txt")
```
